**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**

### 2021-02-03 Manuscripts/Manuscript_AM_v2.Rmd
Updated manuscript and references

 The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of more complex models beyond LMEMs. In particular, $\textsf{R}$ has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g. hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user. At the same time, $\textsf{R}$ has many tools that simplify data simulation, an emerging strategy used to test statistical models [@haverkamp2017]. Data simulation methods allow the researcher to create and  explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and can be also used for power calculations and study design questions.

@@

## The "frequentist" case for longitudinal data

@@

In a biomedical longitudinal study, two or more groups of subjects (patients, mice, samples) are subject to different treatments (e.g., a "treatment" group receives a novel drug vs. a "control" group that receives a placebo), and measurements from each subject within each group are collected at specific time points. The collected response has both  _fixed_ and  _random_ components. The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e, the effect of the novel drug in the "treatment" group. The _random_ component can be defined as "noise" caused by some inherent variability within the study. For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while others are measured in the afternoon, it is possible that the difference in the collection time of the measurement introduces some "noise" in the signal. As their name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.

@@

\begin{equation}
y_{ijt} = \mu+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\pi_{ij} +\varepsilon_{tij}\\ 
(\#eq:linear-model)
\end{equation}

In this model $y_{ijt}$ is the response by subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\mu$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$. A _random effect_ $\pi_{ij}$ of each subject within each group, and independent errors per subject per time per group $\varepsilon_{tij}$, that represent random variation not explained by the _fixed_ or _random_ effects. 
Suppose two treatments groups are used in the study. Then the models per group can be written as below with $j=0$ representing the first treatment group (Group A) and $j=1$ representing the second treatment group (Group B). The linear models then become:

\begin{equation}
y_{ijt} = \begin{cases}
\mu + \beta_1\times time_{t}+\varepsilon_{ijt}+\pi_{ijt}   & \mbox{if Group A}\\
\mu + \beta_1 \times time_{t} + \beta_2 \times treatment_{j} +\beta_3 \times time_{t} \times treatment_{j}+\pi_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
\end{cases}
(\#eq:ANOVA-by-group)
\end{equation}

To further simplify the expression, let us substitute $\widetilde{\mu}=\mu+\beta_{2}$ and $\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}$ in the equation for Group B. The equation then becomes:

\begin{equation}
y_{ijt} = \begin{cases}
\mu + \beta_1\times time_{t}+\varepsilon_{ijt}+\pi_{ijt}   & \mbox{if Group A}\\
\widetilde{\mu} + \widetilde{\beta_1} \times time_{t} +\pi_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
\end{cases}
(\#eq:ANOVA-lines)
\end{equation}

Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different slopes per group). In other words, When it is stated that an rm-ANOVA model "expects" a linear relationship between the covariates and the response, this means that either presented as \@ref(eq:linear-model), \@ref(eq:ANOVA-by-group) or \@ref(eq:ANOVA-lines), an rm-ANOVA model is only able to accommodate linear patterns in the data, the maximum flexibility being that it can assign different slopes to each group. Because the construction of the model is essentially the same in LMEMs, the expected linearity holds in such models, although it is possible to fit models with different slopes at the subject-level[@pinheiro2006].  
When the data does not follow a linear trend, the fit that an rm-ANOVA or LMEM model produces does not accurately represent the trends in the data. To demonstrate this, longitudinal data where a normally distributed response of two groups of 10 subjects each is simulated, and a rm-ANOVA model  (\@ref(eq:linear-model)) is fitted to the data using the package _nlme_ (the code for the simulated data and models is available in the Appendix).

>>>>

The current advances in programming languages designed for statistical analysis (specifically $\textsf{R}$), have eased the computational implementation of more complex models beyond LMEMs. In particular, $\textsf{R}$[@r]has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g. hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user. At the same time, $\textsf{R}$ has many tools that simplify data simulation, an emerging strategy used to test statistical models [@haverkamp2017]. Data simulation methods allow the researcher to create and  explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and can be also used for power calculations and study design questions.

@@

## The repeated measures ANOVA

@@

In a biomedical longitudinal study, two or more groups of subjects (patients, mice, samples) are subject to different treatments (e.g., a "treatment" group receives a novel drug vs. a "control" group that receives a placebo), and measurements from each subject within each group are collected at specific time points. The collected respnse is modeled with both  _fixed_ and  _random_ components. The _fixed_ component can be understood as a constant value in the response which the researcher is interested in measuring, i.e, the average effect of the novel drug in the "treatment" group. The _random_ component can be defined as "noise" caused by some inherent variability within the study. For example, if the blood concentration of the drug is measured in certain subjects in the early hours of the morning while others are measured in the afternoon, it is possible that the difference in the collection time of the measurement introduces some "noise" in the signal. As their name suggests, this "random" variability needs to be modeled as a variable rather than as a constant value.

@@

\begin{equation}
y_{ijt} = \beta_0+ \beta_1 \times time_{t} + \beta_2 \times treatment_{j} + \beta_3 \times time_{t}\times treatment_{j}+\mu_{ij} +\varepsilon_{tij}\\ 
(\#eq:linear-model)
\end{equation}

In this model $y_{ijt}$ is the response by subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively. The _random effect_ $\mu_{ij}$, accounts for variability in each subject within each group by allowing different values for each subject. Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ or _random_ effects. 
In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs. "novel drug" or "saline" vs. "chemotherapy"). Then, the group terms in Equation \@ref(eq:linear-model) can be written as below with $j=0$ representing the first treatment group (Group A) and $j=1$ representing the second treatment group (Group B). The linear models then can be expressed as

\begin{equation}
y_{ijt} = \begin{cases}
\beta_0 + \beta_1\times time_{t}+\mu_{ijt}+\varepsilon_{ijt}   & \mbox{if Group A}\\
\beta_0 + \beta_1 \times time_{t} + \beta_2 \times treatment_{j} +\beta_3 \times time_{t} \times treatment_{j}+\mu_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
\end{cases}
(\#eq:ANOVA-by-group)
\end{equation}

To further simplify the expression, substitute $\widetilde{\mu}=\beta_0+\beta_{2}$ and $\widetilde{\beta_{1}}=\beta_{1}+\beta_{3}$ in the equation for Group B. This substitution allows for a different intercept and slope for Group B. The model is written as

\begin{equation}
y_{ijt} = \begin{cases}
\beta_0 + \beta_1\times time_{t}+\mu_{ijt}+\varepsilon_{ijt}   & \mbox{if Group A}\\
\widetilde{\mu} + \widetilde{\beta_1} \times time_{t} +\mu_{ijt}+\varepsilon_{ijt}  & \mbox{if Group B}\\
\end{cases}
(\#eq:ANOVA-lines)
\end{equation}

Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group). In other words, When it is stated that an rm-ANOVA model "expects" a linear relationship between the covariates and the response, this means that either presented as Equation \@ref(eq:linear-model), Equation \@ref(eq:ANOVA-by-group) or Equation \@ref(eq:ANOVA-lines), an rm-ANOVA model is only able to accommodate linear patterns in the data. If the data show non-linear behavior, the rm-ANOVa model will approximate this behavior with non-parallel lines. Because the construction of the model is essentially the same in LMEMs, the linearity assumption holds for these models, although it is possible to fit models with different slopes at the subject-level[@pinheiro2006].

When the data does not follow a linear trend, the fit that an rm-ANOVA or LMEM produces may not accurately represent the trends in the data. To demonstrate this inaccurate representation, longitudinal data where a normally distributed response of two groups of 10 subjects each is simulated, and a rm-ANOVA model  (Equation \@ref(eq:linear-model)) is fitted to the data using $\textsf{R}$[@r] and the package _nlme_[@nlme] (code in the Appendix). Briefly, the mean responses for each group are linear functions with different intercepts.
